The role of aneuploidy in the evolution of cancer drug resistance

### 2023-12-31 EvolutionaryRescue1.tex
Add files via upload
Remus updates 31 December

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes and observe that aneuploidy plays an important role in the survival of the cancer cell population when the initial tumor size is not too large and the effect of the anti-cancer drug on the aneuploid cells is not too severe.
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely~\citep{komarova2004optimal}.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.

Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time. An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent. Evolutionary rescue in a quasi-periodic environment has been investigated by \citet{marrec2023evolutionary} with the observation that stochasticity increases the chance of extinction in comparison to the case with deterministic environmental fluctuations. Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute. Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fischer's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$. We assume that both chromosomal missegregation and mutation occur during the process of mitosis. Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively, which we assume is lower than the division rates, $v\cdot\max{(\lambda_w, \lambda_a)}<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$.

%Thus, the changes in the number of each cell type is described by 
%\begin{equation}
%\begin{aligned}
%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
%m_t&\rightarrow m_t-1:\quad \mu_am_t.
%\end{aligned}
%\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Stochastic simulations} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (left column, see \Cref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log10}{\log\left(\lambda_m / \mu_m\right)}\right\rfloor + 1.
\end{equation*}

@@

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= \lambda_a,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
\lambda_m' &= \lambda_m,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
\end{align*}
where $K$ is the maximum carrying capacity. The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$. % TODO what constants? see Remus version. Ok

@@

% TODO: OVERVIEW OF RESULTS  ok

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$. We observe that aneuploidy has a significant contribution towards the tumor overcoming chemotherapeutic drugs. 
% TODO PLUG IN REAL VALUES ok

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells \citep{wang2019molecular,mason2017functional}), which are likely to have a deleterious effect in the absence of the drug, $s$ \citep{replogle2020aneuploidy,giam2015aneuploidy}. % TODO refs OK
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}

Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$). Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.

For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=5\cdot10^{-4}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx0.125$ which means that standing genetic variation is the main driver of adaptation even when a relatively small fraction cancer cells are aneuploid in the drug-free environment (in this case $f=20\%$). Furthermore, most tumors display significantly higher levels of aneuploidy such that $f\approx 95\%$ even before cancer treatment. In this case, the population threshold ratio becomes $\tilde{N}^*_a/N^*_a\approx0.0025$ and standing genetic variation is a very significant driver of adaptation to the chemotherapeutic drug.
% TODO PLUG IN REAL VALUES ok

>>>>

\allowdisplaybreaks

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes. Additionally we obtain the mean recurrence time for the tumor to reach its initial size. We observe that aneuploidy plays an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding what factors contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}, which is caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy. Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then exposed to chemotherapeutic agents such as vemurafenib. When compared to wildtype cancer cell the reversine treated cells had higher proliferation in the environment altered by anti-cancer drugs. Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype cell even while they had lower fitness before the onset of chemotherapy. A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is bu antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}

 An important aspect of aneuploidy is the rate with which cell acquire it, which is several orders of magnitude higher when compare to mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of missegration in order to combat cancer \citep{lee2016effects}.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.

\citet{gunnarsson2020understanding} analyze a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue. Evolutionary rescue in one step in which an initially declining population, after a sudden environmental change, has to acquire a mutation has been studied in the context of population genetics  by \citep{orr2008population,orr2014population}. They analyzed a model where the mutant strain is present in small number at the onset of therapy and concluded that this can significantly enhance the chance that the population will survive.

Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.  Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

Here we build on this previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. Besides the probability of evolutionary rescue we calculate the calculate the mean time that an initially declining tumor cell population reaches the initial population size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$. We assume that both chromosomal missegregation and mutation occur during the process of mitosis. Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$. We assume that wildtype cells acquire aneuploidy before the onset of therapy with rate $\tilde{u}\lambda_w$ and that aneuploid cells have a fitness cost $s$ in the drug-free environment.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Stochastic simulations} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (left column, see \Cref{figureAneuploidy}):

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log10}{\log\left(\lambda_m / \mu_m\right)}\right\rfloor + 1.
\end{equation*}

@@

\paragraph{Parameterization.}
To parametrize the model we assume that the cells under consideration are melanoma cells and and rely on \citep{rew2000cell} and \citep{bozic2013evolutionary} for the division and death rates, respectively. \citet{rew2000cell} report in vivo measurements of the potential doubling times (which is the time which the tumor cells would take to double in number if cell loss was not present) for a large set of cancer types.  The division rate is obtained as $\lambda=\log2/T\approx0.1\text{ days}^{-1}$ We select this to be the division rate for wildtype and mutant cells. %We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.

\citet{bozic2013evolutionary} report the growth rate $\Delta_w$ for the wildtype melanoma cancer cells from which they deduce the death rate $\mu_w$ to in the interval $\left[0.11,0.17\right]$ and we choose $\mu_w=0.14days^{-1}$. Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01 which we select as the growth rate of mutant cells as we assume that the mutation makes the cancer cells immune to the chemotherapeutic agent. As a result, we choose $\mu_m=0.1-0.01=0.09\text{ days}^{-1}$ as the death rate for mutant cells.

 Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09\text{ days}^{-1}$, given that aneuploidy increases resistance to chemotherapeutics, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}. Aneuploid birth rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $\lambda_a$ belong to the interval $\left[0.06,0.1\right]$. For most of the paper we will use $\lambda_a=0.0899\text{ days}^{-1}$ such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant which rescues the cancer cell population.

We set the mutation rate $v$ to be equal to $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers immunity to chemotherapy we select $v=10^{-7}$ per cell division. The missegregation rate is chosen from \citet{bakker2023predicting} who determined \textit{in silico} that it must be in the interval $10^{-3}-10^{-2}$ per chromosome per cell division. Given that only about one to two chromosome additions are ``beneficial'' for cancer resistance (see details next paragraph) we choose $u=10^{-2}$ per cell division. For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ (i.e.  $\tilde{u}=10^{-3}$ per cell division) as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improved the fitness of cancer cells in an environment altered by chemotherapeutic drugs. We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. \citet{ippolito2021gene} observed that trisomy 2 and 6 are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a cancer is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the percentage of cancer cells with favorable aneuploidy as $2/22\times0.15\approx 0.0136$ (i.e. $1.36\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy) which we select as the value for $f$ (this gives us a fitness cost of $s= 7.3\times10^{-3}\text{ days}^{-1}$). We note that when we refer to wildtype cancer cells we include those cells which have any aneuploidy except trisomy 2 and 6 as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by therapeutics such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources. This is a good approximation as we assume that the drug cause the wildtype population to decline rapidly and, as a result, density effect can be neglected.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= \lambda_a,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
\lambda_m' &= \lambda_m,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
\end{align*}
where $K$ is the maximum carrying capacity. The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$.

@@

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$. We observe that aneuploidy has a significant contribution towards the tumor overcoming chemotherapeutic drugs. 
% TODO PLUG IN REAL VALUES ok

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells \citep{wang2019molecular,mason2017functional}), which are likely to have a deleterious effect in the absence of the drug, $s$ \citep{replogle2020aneuploidy,giam2015aneuploidy}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}

Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$). Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).

For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx1.825$ which means that de-novo genetic variation is  the main driver of adaptation.
